---
title: "Finding an Edge in Fantasy Football"
output: html_document
---


```{r set-options, echo=FALSE, cache=FALSE}
options(width = 10000)
```

Statistics have become popular for fantasy football but it's unclear if analytics are necessarily better than just using the eye test, or gut feeling, etc. In this post  I'll compare ADP to outside projections to see where ADP has inefficiencies and where there might be edges in projecting players. I was able to get projections from 2012-present from [fantasyfootballanalytics.net](http://apps.fantasyfootballanalytics.net/projections) and from [fftoday.com](fftoday.com/rankings/playerproj.php?Season=201808&PosID=30&LeagueID=1), and ADP data from [ffcalculator.com](https://fantasyfootballcalculator.com/adp). 
I also have my own aggregated projections. ffanalytics.net is an aggregated projection source where they tried to maximize accuracy and so I think it will be a pretty good source to use. All together I have a dataframe like this:



```{r include=F, eval=T , echo=F}
getwd()
library(gridExtra)
library(knitr)
library(plyr)
library(dplyr)
library(reshape2)
library(ggplot2)
myProj<-read.csv("Projections_HALF.csv")
load("Draft Data.RData")
```

```{r include=T, eval=T , echo=F}
myProj<-myProj[grepl("WR|QB|TE|DST|RB|K", myProj$PosFFA), c("Player", "Season", "PosFFA", "fantPts_agg", "fantPts_FDATA", "fantPts_FFTODAY", "fantPts_FFA", "fantPts", "DepthTeam")]
myProj<-myProj[!(myProj$Player=="Michael Thomas"& myProj$DepthTeam!="Nor"),]
myProj$DepthTeam<-NULL

adp<-merge(ffcalc[ffcalc$Year%in% 2012:2018,], myProj[myProj$Season%in% 2012:2018,],
           by.x=c("Player", "Year", "Pos"), by.y=c("Player", "Season", "PosFFA"), all.x=T)
adp<-adp[!is.na(adp$fantPts_agg), ]
adp<-adp[order(adp$ADP_half, decreasing=F),]
kable(adp[adp$Pos%in% c("WR", "QB", "TE", "RB", "DST")& adp$Year==2017,c("Player", "Year", "ADP_half", "fantPts_FFTODAY", "fantPts_FFA","fantPts_agg", "fantPts")][1:5,])
```


Note: all projections are Half-PPR Yahoo Scoring. Next, I'll plot the accuracy of the projections for players who were drafted:

```{r include=T, eval=T , echo=F, fig.width=8.5, fig.height=4, fig.align="center"}

adp<-ddply(adp, .(Pos, Year), mutate,
           ADP.Rank=rank(ADP_half),
           FFA.Rank=rank(-fantPts_FFA),
           FFTODAY.Rank=rank(-fantPts_FFTODAY),
           # FDATA.Rank=rank(-fantPts_FDATA),
           Actual.Rank=rank(-fantPts),
           My.Rank=rank(-fantPts_agg))
adp<-adp[order(adp$ADP_half, decreasing = F), ]
# summary(adp$fantPts[adp$Year!=2018])

# adp[adp$Year==2017& adp$Pos=="K",][1:10,]

corDF<-ddply(adp[adp$Year!=2018,], .(Pos), summarize,
             n=length(fantPts),
             ADP.Rank=cor(Actual.Rank, ADP.Rank),
             FFTODAY.Rank=cor(Actual.Rank, FFTODAY.Rank),
             FFA.Rank=cor(Actual.Rank, FFA.Rank),
             # FDATA.Rank=cor(Actual.Rank, FDATA.Rank),
             My.Rank=cor(Actual.Rank, My.Rank) )
corDF<-melt(corDF, id.vars= "Pos", value.name ="COR",variable.name = "Variable" )

#store sample sizes
n<-corDF[corDF$Variable=="n",]
corDF<-corDF[corDF$Variable!="n", ]

#no projections for these variables
corDF<-corDF[!(corDF$Pos=="K"& corDF$Variable%in% c("FDATA.Rank", "FFTODAY.Rank")),]
corDF<-corDF[!(corDF$Pos=="DST"& corDF$Variable=="FFA.Rank"),]

ggplot(corDF, aes(factor(Pos), COR, fill = Variable)) + 
  geom_bar(stat="identity", position = "dodge") + 
  scale_fill_brewer(palette = "Set1")+
  labs(title="Correlations to End of Season Rank, 2012-17" ,
       caption= paste0("Sample Sizes: ",paste(n$Pos, n$COR, sep=":", collapse=", ")),
       x = "Pos", y = "COR")   +
  geom_text(aes(label = round(COR, 2)), vjust = -0.5,  position = position_dodge(0.9), size=3)



```


The above plot shows how third-party ranks have performed relative to players' ADP rank. Projections have outperformed draft order for WR/TE/QB, but other positions like RB have been harder to gain an edge. However, while it does look like there is inefficiency in things like WR/TE ADP, anyone who has played fantasy knows that ADP in later rounds is often guided by who has higher upside, and so ADP rank is not exactly a ranking of who will finish highest. To account for this, I'd like to see the effect if I just limit it to top 12 DST/K/QB/TE, top 24 RB, and top 36 WR (ranked by ADP), AKA the starting players, so I do that below:

```{r include=T, eval=T , echo=F, fig.width=8.5, fig.height=4.5, fig.align="center"}

adp$ADP.Starter<-ifelse(adp$Pos=="RB"& adp$ADP.Rank<=24, 1, 
                        ifelse(adp$Pos%in% c("QB", "DST", "K", "TE")  &   adp$ADP.Rank<=12, 1, ifelse(adp$Pos=='WR' & adp$ADP.Rank<=36, 1, 0)))

corDF<-ddply(adp[adp$Year!=2018& adp$ADP.Starter==1,], .(Pos), summarize,
             n=length(fantPts),
             ADP.Rank=cor(Actual.Rank, ADP.Rank),
             FFTODAY.Rank=cor(Actual.Rank, FFTODAY.Rank),
             FFA.Rank=cor(Actual.Rank, FFA.Rank),
             # FDATA.Rank=cor(Actual.Rank, FDATA.Rank),
             My.Rank=cor(Actual.Rank, My.Rank) )
corDF<-melt(corDF, id.vars= "Pos", value.name ="COR",variable.name = "Variable" )

#store sample sizes
n<-corDF[corDF$Variable=="n",]
corDF<-corDF[corDF$Variable!="n", ]


#no projections for these variables
corDF<-corDF[!(corDF$Pos=="K"& corDF$Variable%in% c("FDATA.Rank", "FFTODAY.Rank")),]
corDF<-corDF[!(corDF$Pos=="DST"& corDF$Variable=="FFA.Rank"),]

ggplot(corDF, aes(factor(Pos), COR, fill = Variable)) + 
  geom_bar(stat="identity", position = "dodge") + 
  scale_fill_brewer(palette = "Set1")+
  labs(title="Correlations to End of Season Rank, 2012-17, Starters Only" ,
       caption= paste0("Sample Sizes: ",paste(n$Pos, n$COR, sep=":", collapse=", ")),
       x = "Pos", y = "COR")   +
  geom_text(aes(label = round(COR, 2)), vjust = -0.5,  position = position_dodge(0.9), size=3)

```


The above plot shows similar trends to the first. I should mention that for these plots, I am converting the outside projections into ranks to make it more comparable-- the results if I didn't do this were similar so I just left those plots off. From these above plots, I have the following takeways: 

* DST=impossible to project and a waste of a pick.
* K=worth a pick since easier to predict, ADP is similarly accurate as projections
* RB=ADP is similarly as accurate as any projection source, so not much edge. Also, hard to predict
* QB= hard to predict, small edge might be possible
* WR/TE= easiest to predict and edge appears possible using projections

Overall, ADP i.e. consensus opinion is pretty accurate for positions like RB. On the other hand, WR/TE's benefit the most from taking an analytical approach.
Below I show some more plots..Here I examinine the WR projections further to see how they have looked compared to ADP Rankings:



```{r include=T, eval=T , echo=F, fig.width=14, fig.height=8, fig.align="center"}
makePlot<-function(y, cex=.75){
  vars<-c( "My.Rank", "My.Rank")
  bool<-adp$Pos=="WR"& !is.na(adp$My.Rank)  &   (adp$My.Rank%in%1:36|adp$ADP.Rank%in%1:36) 
  plot(adp$ADP.Rank[bool& adp$Year==y]~(rowMeans(adp[bool& adp$Year==y, vars])), xlab="My.Rank", ylab="ADP.Rank (<-Overvalued, Undervalued->)", main=paste0("My.Rank vs ADP.Rank, ",y ," WRs"))
  text( (rowMeans(adp[bool& adp$Year==y, vars])), adp$ADP.Rank[bool& adp$Year==y], labels = adp$Player[bool& adp$Year==y], col='red', pos = 2, cex = cex)
  if(y<=2017){
    text( (rowMeans(adp[bool& adp$Year==y, vars])), adp$ADP.Rank[bool& adp$Year==y], 
          labels = sprintf("%+g",  adp$ADP.Rank[bool& adp$Year==y]-adp$Actual.Rank[bool& adp$Year==y]), col='blue', pos = 4, cex = cex)
  }
  lines(1:100, 1:100, col="blue", lty=2)
  # lines(1:100, 1:100+7, col="purple", lty=2)
  
}
par(mfrow=c(1,2))
makePlot(2016, cex=1)
makePlot(2017, cex=1)

```


The above plots show how projections were able to distinguish players who were undervalued/overvalued. The blue label shows how much they overperformed their ADP. For example, in both years, I had Jarvis Landry as undervalued relative to his ADP and in both years he overperformed his ADP. Obbiously, there are hits and misses, but for the most part it appears that there are much more hits than misses, which is expected based on the higher correlation. Finally, the plot for 2018:


```{r include=T, eval=T , echo=F, fig.width=7, fig.height=5, fig.align="center"}

par(mfrow=c(1,1))
makePlot(2018)
```


## Conclusion 
In this post I examined where there might be edges in fantasy football. I found that ADP is hard to beat, but an analytical approach for WR/TE's allows you to gain an edge compared to ADP, while RBs have had less success with analytical approaches. Some guys the stats like relative to their ADP this year include Tyreek Hill, Brandin Cooks, Jamison Crowder, Nelson Agholor, and others. In my next post I'll show the complete projections, but if you want to you can get the data [here](https://github.com/dlm1223/ncaa-preseason/blob/master/misc%20preseason%20conf%20awards.R).

</br>
[View All Posts->](https://www.brianlefevre.com/archives)
<br /><br /></br>